Experimentally Informed Site-Specific Substitution Models Substantially Deepen Viral Divergence Estimates

### 2018-01-18 manuscript/manuscript.tex
tweaks to figures

\newcommand\skhcomment[1]{{\color{magenta}#1}}
\newcommand\jdbcomment[1]{{\color{red}#1}}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.55\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
\jdbcomment{This figure is too big. Maybe put the three panels side-by-side, increase the space between panels, and increase the font sizes. The logo stacks are kind of tall relative to their width.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
from \ref{tab:wsn_low_params}
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.75\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\jdbcomment{I still think text could be bigger. Logo stacks still show trimming, maybe a bit tall? The Model key could probably be to the right or left.}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller. 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figures/tree_doud}}
\caption{\label{fig:tree_doud}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

@@

@@

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
\caption{\label{suppfig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
}
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
\caption{\label{suppfig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
}
\end{suppfig}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_average}}
\caption{\label{fig:prefs_average}
\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.82$)}}
\end{figure}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_lee}}
\caption{\label{fig:experiment_lee}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} The phylogenetic tree of the "high" alignment. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}

@@

\begin{table}[t!]
\caption{\label{tab:wsn_low_params}
Model parameters used in  Fig.~\ref{fig:decay}.}
      \begin{tabular}{ccccc}
        \hline
          Model & Parameters\\ \hline
          ExpCM & $\beta=1.54196$\\
           & $\kappa=3.47184$\\
           & $\omega=0.219225$\\ 
           & $\phi_A$=0.386529\\ 
           & $\phi_C$=0.17187\\
           & $\phi_T$=0.230319\\
           & $\pi_{r, A\left(X\right)}$:Doud\\
          YNGKP M0 & $\kappa=2.9984$\\
          & $\omega=0.09076$\\
          & $\phi_{1A}=0.326007$\\
          & $\phi_{1C}=0.136307$\\ 
          &$\phi_{1G}=0.278377$\\
          &$\phi_{2A}=0.382017$\\ 
          &$\phi_{2C}=0.179763$\\ 
          &$\phi_{2G}=0.194067$\\
          &$\phi_{3A}=0.354364$\\
          &$\phi_{3C}=0.190112$\\
          &$\phi_{3G}=0.214605$\\
          YNGKP M5 & $\kappa=2.9984$\\
          & $\omega=0.09076$\\
          & $\phi_{1A}=0.326007$\\
          & $\phi_{1C}=0.136307$\\ 
          &$\phi_{1G}=0.278377$\\
          &$\phi_{2A}=0.382017$\\ 
          &$\phi_{2C}=0.179763$\\ 
          &$\phi_{2G}=0.194067$\\
          &$\phi_{3A}=0.354364$\\
          &$\phi_{3C}=0.190112$\\
          &$\phi_{3G}=0.214605$\\
      \end{tabular}
\end{table}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_lee}}
\caption{\label{fig:tree_lee}
\textbf{Trees optimized with an ExpCM defined by H3 preferences lengthen branches from the focal H3 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H3 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H3 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H3 focal sequence (black circle).
}
\end{figure}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_average}}
\caption{\label{fig:tree_average}
\textbf{Trees optimized with an ExpCM defined by the average of H1 and H3 preferences lengthen branches from both the focal H3 sequence and the focal H1 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by the average preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by the average preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006.
The black triangle indicates the H1 focal sequence and the black circle indicates the focal sequence.
}
\end{figure}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/experiment_doud}}
\caption{\label{fig:experiment_doud}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H1 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} A phylogenetic tree of the "high" alignment was constructed using the GTR-CAT model. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H1 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}

>>>>

\newcommand\skhcomment[1]{{\color{cyan}#1}}
\newcommand\jdbcomment[1]{{\color{red}#1}}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
\jdbcomment{This figure is too big. Maybe put the three panels side-by-side, increase the space between panels, and increase the font sizes. The logo stacks are kind of tall relative to their width.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
from \ref{tab:wsn_low_params}
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\jdbcomment{I still think text could be bigger.}
\skhcomment{Should I remove the second from the left site?}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller. 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/empirical_trees}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

\begin{figure}[H]
\centerline{\includegraphics[width=0.70\textwidth]{figures/empirical_trees_2}}
\caption{\label{fig:empirical_trees2}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/doud_compete}}
\caption{\label{fig:doud_compete}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} The phylogenetic tree of the "high" alignment. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}

@@

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
\caption{\label{suppfig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.54$)} 
}
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
\caption{\label{suppfig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.76$)} 
}
\end{suppfig}

\begin{figure}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
\caption{\label{fig:prefs_average}
\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.77$)}}
\end{figure}

@@

\begin{table}[t!]
\caption{\label{tab:wsn_low_params}
Model parameters used in  Fig.~\ref{fig:decay}.}
      \begin{tabular}{ccccc}
        \hline
          Model & Parameters\\ \hline
          ExpCM & $\beta=1.54196$\\
           & $\kappa=3.47184$\\
           & $\omega=0.219225$\\ 
          YNGKP M0 & $\kappa=2.9984$\\
          & $\omega=0.09076$\\
          YNGKP M5 & $\kappa=2.9984$\\
          & $\omega=0.09076$\\
      \end{tabular}
\end{table}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/lee_competing}}
\caption{\label{fig:lee_compete}
\textbf{The ExpCM defined by H1 preferences lengthen longer branches on the HA tree.} 
\textbf{(A)} An HA alignment was subsampled to create three smaller alignments with varying degrees of divergence from the focal H3 sequence, referred to as "low", "intermediate", and "high". 
\textbf{(B)} The phylogenetic tree of the "high" alignment. 
The colors denote the alignment and the black circle denotes the focal H3 sequence. 
\textbf{(C)} The value of the ExpCM and ExpCM+$\Gamma\omega$ stringency parameter $\beta$ decreases as the divergence from the focal H3 sequence increases. 
\textbf{(D)} Comparisons of branch lengths optimized by the four substitution models for the varying degrees of divergence. 
Black points represent branches from the focal H3 sequence and grey points represent all other branches.  
The branch lengths are in average number of codon substitutions per site. 
}
\end{figure}
